---
output:
  github_document:
    html_preview: true
---

<!-- 
This file is licensed with the Creative Commons Attribution 4.0 International License.
-->


```{r setup, include = FALSE}
## get R packages
library(weathercan) #weather data from Environment Canada
library(dplyr) #data munging
library(ggplot2) #plotting
library(sf) #mapping
library(bcmaps) #get bc maps
#library(gganimate) #animate plots 
library(patchwork) #side by side plots (package available from GitHub, https://github.com/thomasp85/patchwork)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE)

theme_plots <- labs(x = NULL, y = "wind speed (km/h)") +
  geom_hline(yintercept = 50, linetype = 2) +
  scale_y_continuous(limits = c(0, 60),
                     breaks = seq(0, 60, 10),
                     expand = c(0, 0))
```


## Weather with `weathercan` `r emo::ji("canada")`

Victoria, British Columbia&mdash;where I live&mdash;has had a bit of [wintery weather](https://www.cheknews.ca/victoria-had-more-february-snow-than-much-of-canada-535410/) this year. 

And.

This happened `r emo::ji("scream")`

<img src = "images/fence.jpg" width = "500"></img>

To be fair, [Environment and Climate Change Canada's](https://weather.gc.ca/city/pages/bc-85_metric_e.html) special weather warning that day did say winds could damage buildings and structures. But seriously, how windy does it need to be to blow down my 2-year-old fence?

I decided to turn this mini-homeowner drama into a reason to try out the [rOpenSci](https://ropensci.org/) [`weathercan` R package](http://ropensci.github.io/weathercan/) `r emo::ji("package")`. `weathercan` makes it easier to search for and download weather data from the [Environment and Climate Change Canada (ECCC) website](http://climate.weather.gc.ca/historical_data/search_historic_data_e.html).

Let's find the nearby ECCC weather stations using `weathercan::station_search()`:

```{r allstns}
library(weathercan)

stns <- stations_search(name = "Esquimalt")
stns
```

Looks like 3 stations for Esquimalt in the station list, however only one&mdash;`ESQUIMALT HARBOUR (52)`&mdash;is currently collecting hourly data. 

```{r stn52}
stns %>% filter(end == 2018 & interval == "hour")
```

A quick map with the `bcmaps` & `sf` R packages confirms the station (red point) is actually in Esquimalt:

```{r stn}
#municipalities in the Capital Regional District
crd <- c("Central Saanich", "Colwood", "Esquimalt", "Highlands",
         "Langford", "Metchosin", "North Saanich", "Oak Bay", "Saanich",
         "Sidney", "Sooke", "Victoria", "View Royal")

esq <- municipalities() %>% 
  filter(ADMIN_AREA_ABBREVIATION == "Esquimalt")

bcmaps::municipalities() %>% 
  filter(ADMIN_AREA_ABBREVIATION %in% crd) %>% 
  ggplot() +
  geom_sf() +
  geom_sf(data = esq, fill = "grey10") +
  geom_sf(data = st_as_sf(stn52, coords = c("lon", "lat"),
                          crs = 4326),
  size = 4, colour = "red", show.legend = TRUE) +
  coord_sf(datum = NA) +
  labs(title = "Township of Esquimalt & Surrounding\nMunicipalities of the Capital Region District ") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

So let's grab all the data for station #52 `ESQUIMALT HARBOUR` using `weathercan::weather_dl()` and have a look at the wind speeds:

```{r raw_data}
data <- weather_dl(station_ids = 52)

data %>%
  select(station_id, station_name, date, wind_spd) %>%
  ggplot(aes(x = date, y = wind_spd)) +
  geom_point(colour = "#3182bd", alpha = .7) +
  theme_minimal() +
  labs(x = NULL, y = "avg hourly wind speed (km/h)",
       title = "Average Hourly Wind Speeds Measured at Esquimalt Harbour") +
  geom_hline(yintercept = 50, linetype = 2) +
  scale_y_continuous(limits = c(0, 70),
                     breaks = seq(0, 70, 10),
                     expand = c(0, 0))
```

Well, that was disapointing `r emo::ji("disappointed")`, looks like the recent high winds were _not_ that unique afterall&mdash;lot's of [average wind speeds](http://climate.weather.gc.ca/glossary_e.html#windSpd) in the same range or even higher in the past couple decades. I was hoping for something much more dramatic (read explanatory). As a side note, there looks like there was an instrument change in 2014? `r emo::ji("person_shrugging")`

But I am not ready to give up. The fence was only built in April 2017&mdash;maybe we should look at wind speeds since then?

```{r last2yrs}

p1 <- data %>%
  select(station_id, station_name, date, wind_spd) %>%
  filter(date > "2017-04-01") %>%
  ggplot(aes(x = date, y = wind_spd)) +
  geom_point(colour = "#3182bd", alpha = .7) +
  theme_minimal() +
  labs(x = NULL, y = "avg hourly wind speed (km/h)",
       title = "Wind Speeds Since Fence was Built") +
  geom_hline(yintercept = 50, linetype = 2) +
  scale_y_continuous(limits = c(0, 60),
                     breaks = seq(0, 60, 10),
                     expand = c(0, 0))

p2 <- data %>%
  select(station_id, station_name, date, wind_spd) %>%
  filter(date > "2018-12-10") %>%
  ggplot(aes(x = date, y = wind_spd)) +
  geom_point(colour = "#3182bd", alpha = .7) +
  theme_minimal() +
  labs(x = NULL, y = "avg hourly wind speed (km/h)",
       title = "Wind Speeds During Recent Storms") +
  geom_hline(yintercept = 50, linetype = 2) +
  scale_y_continuous(limits = c(0, 60),
                     breaks = seq(0, 60, 10),
                     expand = c(0, 0))

p1 + p2
```

`r emo::ji("tada")` The recent blast of wintery weather _was_ the first test of the fence afterall. But why did the fence fair OK with the December winds but fall over in February? There is a fair bit of over-plotting with these points, so let's look _even_ closer `r emo::ji("eyes")`:

```{r storm}
p3 <- data %>%
  select(station_id, station_name, date, time, wind_spd) %>%
  filter(time >="2018-12-20 04:00:00" & date < "2018-12-21 04:00:00") %>%
  ggplot(aes(x = time, y = wind_spd)) +
  geom_point(colour = "#3182bd", alpha = .7) +
  theme_minimal() +
  labs(x = NULL, y = "avg hourly wind speed (km/h)",
       title = "December Wind Storm") +
  geom_hline(yintercept = 50, linetype = 2) +
  scale_y_continuous(limits = c(0, 60),
                     breaks = seq(0, 60, 10),
                     expand = c(0, 0))

p4 <- data %>%
  select(station_id, station_name, date, time, wind_spd) %>%
  filter(time >="2019-02-08 14:00:00" & date < "2019-02-10 00:00:00") %>%
  ggplot(aes(x = time, y = wind_spd)) +
  geom_point(colour = "#3182bd", alpha = .7) +
  theme_minimal() +
  labs(x = NULL, y = "avg hourly wind speed (km/h)",
       title = "February Wind Storm") +
  geom_hline(yintercept = 50, linetype = 2) +
  scale_y_continuous(limits = c(0, 60),
                     breaks = seq(0, 60, 10),
                     expand = c(0, 0))

p3 + p4
```

Well one thing we can say about the February storm is that it went on, and on, and on&mdash;high winds recorded for over 12 hours `r emo::ji("dash")`! Maybe that is what did my new fence in? Or, maybe it was not average hourly wind speeds but wind gust speeds? Unfortunately wind gust speed is not available in the [Environment and Climate Change Canada (ECCC) data set](http://climate.weather.gc.ca/historical_data/search_historic_data_e.html). 

Ah well. Maybe fence version 1 was just not meant to be&mdash;version 2 is underway and _will_ be ready for future winter weather `r emo::ji("canada")`.


- The `weathercan` R package is part of the [rOpenSci project](https://ropensci.org/)&mdash;they are transforming science through open data and software `r emo::ji("100")`
- Learn more `weathercan` authored by [Steffi LaZerte](http://steffilazerte.ca/) & Sam Albers [here](http://ropensci.github.io/weathercan/)
- The R source code for this post can be reviewed [here](https://github.com/stephhazlitt/some-assembly-required/blob/master/R/fence/fence.Rmd)


`r Sys.Date()`

