---
output:
  github_document:
    html_preview: true
---

<!-- 
This file is licensed with the Creative Commons Attribution 4.0 International License.
-->


```{r setup, include = FALSE}
## get R packages
library(dplyr)
library(weathercan)
library(ggplot2)
library(envreportutils)
library(magick)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE)

# save(data, file = "tmp/stn52_data.RData")
# load("tmp/stn52_data.RData")
```


## When You Miss the Target `r emo::ji("dart")`

- _Sometimes_ you can do a data analysis and not find anything that really sparks joy&mdash;no correlations, no explanatory patterns `r emo::ji("bar_chart")`

- _Sometimes_ you can do a data analysis and be looking at the wrong data `r emo::ji("mag")`

- _Sometimes_ nice people make suggestions where you might want to look `r emo::ji("eyes")`

In my [short post](https://github.com/stephhazlitt/some-assembly-required/blob/master/R/fence/fence.md) looking at whether average wind speeds (km/h) during the Victoria, British Columbia February 2019 storm might have explained why my fence fell down, I didn't find a "smoking gun" `r emo::ji("gun")`&mdash;the February storm perhaps lasted longer than 'usual' but measured wind speeds were within the normal range of storm wind speeds since 1994. 

A number of people suggested I look at wind direction `r emo::ji("dash")`. So I did. KAPOW `r emo::ji("tada")``r emo::ji("open_mouth")` 

(1) Get and wrangle the [Environment and Climate Change Canada (ECCC)](http://climate.weather.gc.ca/historical_data/search_historic_data_e.html) [wind direction data](http://climate.weather.gc.ca/glossary_e.html#windSpd) using the the [rOpenSci](https://ropensci.org/) [`weathercan` R package](http://ropensci.github.io/weathercan/) `r emo::ji("package")`. I filtered out wind speeds of zero and grouped wind speed into a three categories (1-25 km/h, 26-50 km/h and 51-75 km/h):

```{r wind_data}
library(weathercan)
library(dplyr)
library(ggplot2)

stn <- stations_search(name = "Esquimalt", interval = "hour") %>% 
  select(station_id) %>% 
  pull()

data <- weather_dl(station_id = stn)

wind_dir <- data %>% 
  select(station_id, station_name, date, wind_spd, wind_dir) %>% 
  filter(wind_spd != 0) %>% 
  mutate(wind_spd_cat = cut(wind_spd, breaks = c(0,25,50,75),
                            labels = c("1-25 km/h Winds",
                                       "26-50 km/h Winds",
                                       "51-75 km/h Winds")))
wind_dir
```

(2) Plot wind direction over time for each wind speed category:

```{r wind_plot, fig.width = 9}
wind_plot <- wind_dir %>% 
  filter(wind_dir != 0) %>% 
  mutate(wind_dir_deg = wind_dir*10) %>% 
  ggplot(aes(x = date, y = wind_dir_deg)) +
  geom_point(colour = "#3182bd", alpha = .7) +
  facet_wrap(~wind_spd_cat) +
  theme_minimal() +
  labs(x = NULL, y = "wind direction (degrees true)",
       title = "Wind Direction & Average Hourly Wind Speed Measured\nat Esquimalt Harbour (1994-2019)",
       caption = "*Data sourced from Environment Canada using the weathercan R package") +
  scale_y_continuous(limits = c(-5, 365),
                     breaks = seq(0, 360, 60),
                     expand = c(0, 0)) +
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        strip.text = element_text(size = 10, face = "bold"),
        plot.title = element_text(size = 12, face = "bold"),
        plot.margin = unit(c(4,50,4,4),"mm"))
wind_plot
```

```{r save_img, results = "hide", include = FALSE}
png_retina(filename = "tmp/wind_dir_plot.png",
           width = 700, height = 500)
plot(wind_plot)
dev.off()
```

(3) Get fancy with the `magick` R package `r emo::ji("package")` and overlay my sad fence picture:

```{r final_plot}
library(magick)

wind_plot_img <- image_read("tmp/wind_dir_plot.png")
fence_img <- image_read("images/fence.jpg")

final_img <- image_composite(image_scale(wind_plot_img, "x800"),
                             image_scale(fence_img, "x150"),
                             offset = "+900+550")
final_img <- image_draw(final_img)
rect(860, 540, 1110, 725, border = "red", lty = "dashed", lwd = 3)
final_img
```

```{r save_file, results = "hide", include = FALSE}
image_write(final_img, "tmp/wind_dir_fence.png")
```




`r Sys.Date()`