---
output:
  github_document:
    html_preview: true
---

<!-- 
This file is licensed with the Creative Commons Attribution 4.0 International License.
-->


```{r setup, include = FALSE}
## get R packages
library(glue) #glue strings
library(cansim) #get StatsCan data
library(dplyr) #data munging
library(ggplot2) #plotting
library(scales) #number commas
library(here) #easier paths
library(curl) #pull data from web
library(readxl) #import xls
library(tidyr) #reshape data frame
library(readr) #import csv
library(sf) #mapping
library(bcmaps) #get bc maps
library(gganimate) #animate plots (package available from GitHub, https://github.com/thomasp85/gganimate)
library(patchwork) #side by side plots (package available from GitHub, https://github.com/thomasp85/patchwork)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE)

path <- here("R", "bc-sd61-popn", "data")

theme_plots <- theme(axis.text = element_text(size = 10),
                     plot.title = element_text(size = 14),
                     plot.subtitle = element_text(size = 12),
                     plot.caption = element_text(size = 10),
                     axis.title = element_blank())
```


## The Kids Are Coming, Let's Get Ready

We are in the campaign period for the British Columbia [2018 General Local Elections](https://elections.bc.ca/political-participants/local-elections-campaign-financing/2018-general-local-elections/). I live in the Township of Esquimalt `r emo::ji("cityscape")`&mdash;an amazing community. However, there is always room to improve, even on amazing. I have enjoyed talking with our local government candidates, in particular about what issues motivated them to campaign to step into such important roles&mdash;the housing crisis, responsible development, community building. In turn, I have had a fair bit to say about issues on my mind, such as climate change ([this `r emo::ji("scream")`](https://www.ipcc.ch/news_and_events/pr_181008_P48_spm.shtml)), reducing speed limits in neighbourhood areas `r emo::ji("car")`, more and safer bike paths for family cycling `r emo::ji("bike")`, and, _schools_. Specifically, space in our only catchment elementary school&mdash;[L'Ecole Macaulay Elementary](https://macaulay.sd61.bc.ca/).  

Esquimalt used to have 2 catchment elementary schools, Lampson Street & Macaulay ([Ecole Victor-Brodeur](https://en.wikipedia.org/wiki/%C3%89cole_Victor-Brodeur) is also in Esquimalt, a school that serves the francophone community). Lampson, a school building with [an impressive history in the region](https://victoriadailyphoto.blogspot.com/2013/05/lampson-street-school.html), was closed as a catchment elementary school in 2007 due to declining enrolment. 

However, things have and are changing&mdash;and pretty _fast_. 

Due to the recent, legal change to [revert class size & composition back to 2002 levels](https://bctf.ca/IssuesInEducation.aspx?id=5530), as well as a growing population&mdash;Macaulay has added one or more new classes in each of the 3 years I have had kids in the school, as have many elementary schools throughout the Capital Regional District. Macaulay has over 470 kids enrolled this year, so I have been talking `r emo::ji("speaking_head")` to candidates about how Macaulay is pretty _full_&mdash;and I think we need a plan for the kids that are still to come.

I thought it would be interesting to support this discussion with some _data_. Also, I am always keen to work on my [R](https://www.r-project.org/) skills `r emo::ji("woman_technologist")`. So here it goes.

#### Population Change in Victoria Census Metropolitan Area (2001-2017)

Let's start by pulling some population estimate data from Statistics Canada&mdash;I found data for each Census Metropolitan Area by age group. Perfect. I had to Google to learn about what the Victoria Census Metropolitan Area (CMA) actually includes (`r emo::ji("woman_facepalming")`), turns out it includes [all of the Capital Regional District minus the Gulf Islands](https://www12.statcan.gc.ca/census-recensement/2011/as-sa/fogs-spg/Facts-cma-eng.cfm?LANG=Eng&GK=CMA&GC=935). Let's have a look `r emo::ji("eyes")`:

<details><summary><i>Click to see the R code for the Victoria CMA plots</i></summary>

```{r cma}
## Data: Victoria Census Metropolitan Census Area (British Columbia) Population
## Estimates [Table: 17-10-0078-01 (formerly CANSIM  051-0056) from Statistics Canada

#get Statistics Canada data using cansim package
bc_popn <- get_cansim(1710007801) %>%
  filter(GEO == "Victoria, British Columbia",
         Sex == "Both sexes") %>%
  select(year = REF_DATE, age_group = `Age group`, population_estimate = VALUE) %>% 
  mutate(year = as.integer(year))

#make plot of Victoria CMA population estimates
cma_plot <- bc_popn %>% 
  filter(age_group == "All ages") %>% 
  ggplot(aes(x = year, y = population_estimate, group = 1)) + 
  geom_line(size = 1, colour = "#54278f") +
  scale_x_continuous(limits = c(2001, 2018),
                     breaks = seq(2002, 2017, 3),
                     expand = c(0,0)) +
  scale_y_continuous(limits = c(320000, 380000),
                     breaks = seq(320000, 380000, 5000),
                     expand = c(0,0), labels = comma) +
  labs(title = "Population Estimates for Victoria CMA",
       subtitle = "Include all ages",
       caption = "Data sourced from Statistics Canada") +
  theme_minimal() +
  theme_plots

#kid age groups
kid_ages <- c("0 to 4 years", "5 to 9 years", "10 to 14 years", "15 to 19 years")

#make plot of kid Victoria CMA population estimates
cma_kids_plot <- bc_popn %>% 
  filter(age_group %in% kid_ages) %>% 
  group_by(year) %>% 
  summarise(population_estimate = sum(population_estimate)) %>% 
  ggplot(aes(x = year, y = population_estimate, group = 1)) + 
  geom_line(size = 1, colour = "#54278f") +
  scale_x_continuous(limits = c(2001, 2018),
                     breaks = seq(2002, 2017, 3),
                     expand = c(0,0)) +
  scale_y_continuous(limits = c(64000, 70000),
                     breaks = seq(64000, 70000, 500),
                     expand = c(0,0), labels = comma) +
  labs(title = "Kid Population Estimates for Victoria CMA",
       subtitle = "Kids include ages 0 to 19 years",
       caption = "Data sourced from Statistics Canada") +
  theme_minimal() +
  theme_plots
```

```{r cma-plots, eval = FALSE, fig.width = 12, fig.height = 7}
#plot the Victoria CMA plots
cma_plot + cma_kids_plot
```

</details>

```{r cma-actual-plots, echo = FALSE, fig.width = 12, fig.height = 7}
#plot the Victoria CMA plots
cma_plot + cma_kids_plot
```

The Statistics Canada data confirms what we hear all the time, the population in the Victoria CMA has been steadily increasing since 2001. But _not_ due to more kids&mdash;the population of 0 to 19 year olds steadily declined from 2001 until 2014. This explains the decline in enrolment&mdash;and School District decisions to close schools at that time. However, 2014 brings a clear change&mdash;kid numbers start increasing. But what about in the Township of Esquimalt?

#### Population Change in the Township of Esquimalt (2001-2017)

I found some population data for Municipalities from [BCStats](https://www2.gov.bc.ca/gov/content?id=36D1A7A4BEE248598281824C13CB65B6), but unfortunately not by age group `r emo::ji("disappointed")`.

<details><summary><i>Click to see the R code for the Esquimalt plot & map</i></summary>

```{r mun}
## Data: Esquimalt District Municipality population estimates from
## BCStats Population Estimates webpage

#municipalities I want
crd <- c("Central Saanich", "Colwood", "Esquimalt", "Highlands",
         "Langford", "Metchosin", "North Saanich", "Oak Bay", "Saanich",
         "Sidney", "Sooke", "Victoria", "View Royal")

#get the 2001-2011 CRD population estimate data file 
bcstats_mun_2001_2011 <- 
  "http://www.bcstats.gov.bc.ca/Files/0379a32f-cec8-438d-83e0-6724b2a2a272/BCDevelopmentRegionRegionalDistrictandMuncipalPopulationEstimates2001-2011.xls"

get_mun_2001_2011 <- curl_download(bcstats_mun_2001_2011,
                                   destfile = glue(path, "/bcstats_mun_2001_2011.xls"))

crd_2001_2011 <- read_xls(get_mun_2001_2011, skip = 3) %>%  
  filter(Name %in% crd) %>% 
  select(-`2011`, -Type, -SGC)

# get 2011-2017 CRD population estimate data file and merge data with the 2001-2011 data 
bcstats_mun_2011_2017 <- 
  "http://www.bcstats.gov.bc.ca/Files/285cd56c-9be1-4c5e-a153-3deeffa2ac94/BCDevelopmentRegionRegionalDistrictandMuncipalPopulationEstimates2011-2015.xls"

get_mun_2011_2017 <- curl_download(bcstats_mun_2011_2017,
                                   destfile = glue(path, "/bcstats_mun_2011_2017.xls"))

crd_2001_2017 <- read_xls(get_mun_2011_2017, range = ("A3:J48")) %>% 
  filter(Name %in% crd) %>% 
  bind_cols(crd_2001_2011) %>% 
  select(-`2011`, -`Area Type`, -SGC, -Name1) %>% 
  gather(key =  year, value = population_estimate, -Name) %>% 
  mutate(population_estimate = as.numeric(population_estimate),
         year = as.integer(year)) %>% 
  arrange(year)

#plot of Township of Esquimalt population estimates
esquimalt_plot <- crd_2001_2017 %>% 
  filter(Name == "Esquimalt") %>%
  ggplot(aes(x = year, y = population_estimate, group = 1)) + 
  geom_line(size = 1, colour = "#54278f") +
  labs(title = "Population Estimates for Esquimalt",
       subtitle = "Includes all ages",
       caption = "Data sourced from BCStats") +
   scale_x_continuous(limits = c(2001, 2018),
                     breaks = seq(2002, 2017, 3),
                     expand = c(0,0)) +
  scale_y_continuous(limits = c(16400, 17600),
                     breaks = seq(16400, 17600, 100),
                     expand = c(0,0), labels = comma) +
  theme_minimal() +
  theme_plots

#calculate municipality population change for 2015 to 2017
crd_change <- crd_2001_2017 %>% 
  filter(year %in% c(2015, 2017)) %>%
  group_by(Name) %>% 
  mutate(popchange = population_estimate-lag(population_estimate)) %>% 
  mutate(percchange = round((popchange/lag(population_estimate) * 100), digits = 0)) %>% 
  filter(year == 2017) %>% 
  select(-year) 

#map 2 year % population change for municipalities
crd_change_spatial <- municipalities() %>% 
  filter(ADMIN_AREA_ABBREVIATION %in% crd) %>% 
  mutate(Name = ADMIN_AREA_ABBREVIATION) %>% 
  left_join(crd_change) %>% 
  ggplot() +
  geom_sf(aes(fill = percchange)) +
  coord_sf(datum = NA) +
  labs(title = "CRD Percent Population Change 2015-2017",
       subtitle = "Includes all ages",
       caption = "Data sourced from BCStats") +
  scale_fill_gradient(name = "Percent\nChange", low = "#bcbddc", high = "#132B43") +
  theme_minimal() +
  theme_plots
```

```{r mun-plots, eval = FALSE,  fig.width = 10, fig.height = 7}
#plot the Esquimalt plots
esquimalt_plot + crd_change_spatial
```

</details>

```{r mun-actual-plots, echo = FALSE,  fig.width = 10, fig.height = 7}
#plot the Esquimalt plots
esquimalt_plot + crd_change_spatial
```

The Township of Esquimalt population has been a bit more variable since 2001, with a declining pattern from 2006 to 2013. However, there has been an uptick in recent years. In fact, _all_ the Greater Victoria municipalities&mdash;except North Saanich&mdash;had a measurable increase in popuation over the last 2 years&mdash;Victoria a 1% increase, Esquimalt 3% and Sooke 11%! Wow, that is _a lot_.

What I _really_ want to know is how the population of _kids_ is changing in Esquimalt. I was not able to find population-by-age data for Esquimalt, but I did find it for the Greater Victoria School Districts.

#### Kid Population Change in School District 61-Greater Victoria

<details><summary><i>Click to see the R code for the SD61 plots</i></summary>

```{r sd61}
## Data: School District 61 Population Estimates & Projections from BCStats manual 
## online tool: https://www.bcstats.gov.bc.ca/apps/PopulationProjections.aspx

#get SD61 population estimates & projections
sd61_kids <- read_csv(glue(path, "/Population_Projections.csv")) %>% 
  select(Year, "<1", "1-4", "5-9", "10-14", "15-19") %>% 
  gather(key = age_group, value = population_estimate, -Year) %>% 
  filter(Year < 2038) %>% 
  mutate(age_group = factor(age_group, ordered = TRUE,
                            levels = c("<1", "1-4", "5-9", "10-14", "15-19")),
         data_type = case_when(Year < 2018 ~ "estimated",
                               TRUE ~ "projected"))

#plot of kid population estimates in SD61 by age group
sd61_kids_line_plot <- sd61_kids %>% 
  filter(age_group != "<1") %>% 
  ggplot(aes(Year, population_estimate, group = age_group, colour = data_type)) +
  geom_line(size = 1) +
  geom_segment(aes(xend = 2037 , yend = population_estimate), linetype = 2, colour = 'grey') + 
  geom_point(size = 2) + 
  geom_text(aes(x = 2037.1 , label = age_group), hjust = 0, colour = "#54278f") + 
  coord_cartesian(clip = 'off') + 
  labs(title = "Kid Population Estimates & Projections by Age Group for SD61",
       # subtitle = "Kids include ages 1 to 19 years",
       caption = "Data sourced from BCStats") +
  theme(plot.margin = margin(5.5, 0, 5.5, 5.5)) +
  scale_x_continuous(limits = c(1984, 2040), breaks = seq(1986, 2038, 5), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(6000, 13000, 1000), expand = c(0,0), labels = comma) +
  scale_colour_manual(name = "", values = c("#54278f", "#9e9ac8")) +
  scale_fill_manual(name = "", values = c("#54278f", "#9e9ac8")) +
  theme_minimal() +
  theme_plots +
  theme(legend.position = "top",
        legend.text = element_text(size = 12))

#plot of total sd61 kid population estimates
sd61_kids_total_plot <- sd61_kids %>% 
  group_by(Year, data_type) %>% 
  summarise(population_estimate = sum(population_estimate))  %>% 
  ggplot(aes(x = Year, y = population_estimate, colour = data_type, group = 1)) + 
  geom_point() +
  geom_line(size = 1) +
  labs(title = "Kid Population Estimates & Projections for SD61",
       subtitle = "Kids include ages 0 to 19 years",
       caption = "Data sourced from BCStats") +
  scale_x_continuous(limits = c(1984, 2040), breaks = seq(1986, 2038, 5), expand = c(0,0)) +
  scale_y_continuous(limits = c(35000, 43000), breaks = seq(35000, 43000, 500), expand = c(0,0), labels = comma) +
  scale_colour_manual(name = "", values = c("#54278f", "#9e9ac8")) +
  theme_minimal() +
  theme_plots +
  theme(legend.position = c(.2,.2),
        legend.text = element_text(size = 12)) 
```

```{r sd61-plots, eval = FALSE, fig.width = 7, fig.height = 6, fig.align= "center"}
#animate the sd61 age group plot
sd61_kids_line_plot +
  transition_reveal(age_group, Year)

#plot the sd61 total line plot
sd61_kids_total_plot
```

</details>

```{r sd61-actual-anim, echo = FALSE, fig.width = 8, fig.height = 6, fig.align= "center"}
#animate the sd61 age group plot
sd61_kids_line_plot +
  transition_reveal(age_group, Year)
```

The School District 61 data shows a similar pattern to what we have already observed&mdash; population declines for all kid age groups through the 2000s. We see the recent increases, _and_  the projected increases to come&mdash;for at least the next **15** years! I confess that while the animated plot is fun to make `r emo::ji("nerd")`, rolling the kid age group populations together and keeping things still really tells the story:


```{r sd61-actual-plots, echo = FALSE, fig.width = 8, fig.height = 6, fig.align= "center"}
#plot the sd61 total plot
sd61_kids_total_plot
```


#### Yep, The Kids Are Coming

Our Esquimalt catchment elementary school is full and the kid population estimate lines are going up&mdash;and predicted to do so for at least the next 15 years `r emo::ji("chart_with_upwards_trend")`. In my opinion, a plan for the kids that are still to come should be on the to-do list `r emo::ji("scroll")` for the Township of Esquimalt. The elementary school space crunch has aleady had community impacts throughout the Capital Regional District, displacing dedicated computer labs, after-school care situations, pre-schools. The School District recently launched a [Catchment Boundary Review & Public Consultation](https://www.sd61.bc.ca/news-events/community/catchmentboundaryreview/)&mdash;a great step in supporting planning. However, with only one catchment school and one boundary, I think it is _also_ time to talk about bringing a second catchment elementary school (back) on line for Esquimalt kids `r emo::ji("school")`.

- Details on the data sources (e.g. website url, licences) are [here](https://github.com/stephhazlitt/some-assembly-required/blob/master/R/bc-sd61-popn/data/README.md)
- The R source code can be reviewed [here](https://github.com/stephhazlitt/some-assembly-required/blob/master/R/bc-sd61-popn/sd61_kids.Rmd)
- Learn more about the CRD & Township of Esquimalt candidates  [here](https://www.timescolonist.com/elections)
- Don't forget to get out and [**VOTE on October 20th, 2018**](https://www.esquimalt.ca/municipal-hall/elections/2018-election/information-voters) 

`r Sys.Date()`

